Inference of Bifurcations\\with Differentiable Continuation

### 2021-05-14 docs/article.tex
Updates from Overleaf

\input{style}
\begin{document}
\input{notation}

\title{Inference of Bifurcations with Differentiable Continuation}
\author{Gregory Szep$^{1,2}$, Neil Dalchau$^2$ and Attila Csikasz-Nagy$^{1,3}$}
\affil{$^1$King's College London, UK, $^2$Microsoft Research Cambridge, UK, $^3$Pazmany Peter Catholic University, Budapest, Hungary}
\date{\today} \maketitle

\begin{abstract}
    In this work we propose a gradient-based semi-supervised approach for matching target bifurcations with parameterised differential equation models. The cost function contains a supervised likelihood term that is maximal when predicted bifurcations match the targets, and an unsupervised correction term that encourages bifurcations by maximising the curvature of the determinant of the Jacobian. The calculation of gradients with respect to parameters shares the same computational complexity as deflated pseudo-arclength continuation used to calculate the bifurcation diagram. We demonstrate model synthesis with minimal models which explore the space of saddle-node and pitchfork diagrams, a genetic toggle switch from synthetic biology and the FitzHugh-Nagumo model. Furthermore, the cost landscape allow  us  to  organise  models  in  terms  of topological and geometric equivalence.
\end{abstract}

\section{Introduction}
\subsection{Background}

Backpropagation through differential equation solves has been a breakthrough over the past couple of years \cite{Chen2018NeuralEquations,Rackauckas2019DiffEqFlux.jl-AEquations} that enabled scalable parameter inference for differential equations. Optimisation targets, however, have mostly been expressed in the spatio-temporal domain. Acquisition of such data can be costly and can often contain over-constraining information generated by processing steps or the measurement device rather than the state variables that underpin the observed mechanism. In microscopy, for example, data is often reported in arbitrary fluoresce units allowing the observer to shift and scale data arbitrarily. Furthermore, such data may also not contain sufficient information about dynamical transients in order to identify kinetic parameters. The emerging picture suggests that identification of the qualitative dynamics -- the bifurcation diagram -- should precede any attempt at inferring kinetic parameters \cite{Stumpf2019ParameterBifurcations}. Techniques for back-propagating through implicit equation solvers have also been developed \cite{Look2020DifferentiableLayers,Bai2019DeepModels} although at the time of writing this paper, have not been applied to bifurcation diagrams.

In this work we propose a gradient-based semi-supervised approach that focuses on fitting qualitative dynamics, defined by state space structures, rather than kinetics. Drawing inspiration from implicit solvers \cite{Look2020DifferentiableLayers,Bai2019DeepModels} to calculate gradients we find that their computation shares the same  complexity as the algorithm used to calculate the bifurcation diagram. We use a predictor-corrector method called deflated pseudo-arclength continuation \cite{Farrell2016TheDiagrams,Veltz2019PseudoArcLengthContinuation.jl}, originally developed for partial differential equations. In the case of partial differential equations the computational complexity of calculating a single bifurcation diagram is not bounded since superpositions of localised solutions may give rise to uncountably many branches \cite{Avitabile2010ToEquation}. The complexity of computing a single branch, however, is bounded by the complexities of the chosen eigenvalue solver and corrector, and further decreased by adaptive stepping procedures \cite{Aruliah2016AlgorithmContinuation}.

We find that the cost function landscape contains basins that not only allow us to synthesise models with a desired bifurcation structure but also allow us to organise models in terms of topological and geometric equivalence. We discuss the relevance of this in model selection.

% \subsection{Related Works}
% \begin{itemize}
%     \item smooth and match estimator methods \cite{Ranciati2017BayesianParameters}
%     \item bifurcation inference using Mixed Integer Nonlinear Programming optimisation strategy \cite{Otero-Muras2018Optimization-basedModels}
% \end{itemize}

\clearpage
\section{Method}

\subsection{Problem Statement}
Suppose we parameterise a set of differential equations for states $u\in\Reals^N$ with a function $\rates$ in an unknown parameter space $\theta\in\Reals^M$. We would like these differential equations to obey a set of target bifurcations $\targets:=\{p_1\dots p_K\}$ along a known bifurcation parameter $p\in\Reals$. The following sections outline how a gradient descent algorithm could encourage predicted bifurcations $\predictions(\theta)$ coming from parameterised differential equations to match specified targets $\targets$. This would allow us to design differential equations using high-level qualitative constraints and sample qualitatively equivalent models in regions of optimal $\theta$. Let the differential equations be defined as
\begin{align}
	\partial_t u=\rates(z)
	\qquad\mathrm{where}\quad z:=(u,p)\quad
	\rates : \Reals^{N+1}\rightarrow\Reals^N
	\label{eq:model}
\end{align}

First we view the steady states of this problem in terms of implicit space curves \cite{Goldman2005CurvatureSurfaces} in an augmented state space $z\in\Reals^{N+1}$ which combines the state $u$ and bifurcation parameter $p$. We then realise that the curvature of the Jacobian determinant $\Det(z):=|\partial\rates/\partial u|$ is maximised when bifurcations appear in the observed region of $z$. Luckily the curvature can be evaluated analytically \cite{Goldman2005CurvatureSurfaces}. The steady state regions of $z$ are computed using a parameter continuation method \cite{Veltz2019PseudoArcLengthContinuation.jl,Farrell2016TheDiagrams}. Finally a differentiable semi-supervised cost function $\loss(\theta|\targets)$ is proposed that would allow gradient descent algorithms to efficiently encourage bifurcations and match their locations to targets $\targets$.

For clarity, we guide the reader through the methods with the following minimal models that explore the space of saddle-nodes $\rates(z) = p + \theta_{1}u+\theta_{2}u^3$ and pitchforks $\rates(z) = \theta_{1} + p u+\theta_{2}u^3$. These minimal models are shown with targets $\targets$ in Figure \ref{fig:saddle-node} and Figure \ref{fig:pitchfork} respectively. These figures show that the curvature of the determinant $\Det(z)$ increases in the vicinity of bifurcations and crosses zero at the bifurcations. This gives us our first hint of what should be optimised to encourage bifurcations. Success on these minimal models demonstrates the proof of concept, and success in practice is demonstrated on a genetic toggle switch model taken from synthetic biology and the FitzHugh-Nagumo model.
\begin{figure}[H]
\centering{}
\captionsetup{justification=centering}
\includegraphics[width=11cm]{saddle-node}
\caption{Saddle-node model $\rates(z) = p + \theta_{1}u+\theta_{2}u^3$ with set $\theta=(2,-1)$ and targets $\targets=\{-1/2,1/2\}$. Lighter shades indicate the determinant crossing zero for unstable solutions}
\label{fig:saddle-node}
\end{figure}
\begin{figure}[H]
\centering{}
\captionsetup{justification=centering}
\includegraphics[width=11cm]{pitchfork}
\caption{Pitchfork model $\rates(z) = \theta_{1} + p u+\theta_{2}u^3$ with set $\theta=(-1/10,-1)$ and target $\targets=\{0\}$. Lighter shades indicate the determinant crossing zero for unstable solutions}
\label{fig:pitchfork}
\end{figure}

\subsection{Bifurcation Curves as Tangent Fields}
Suppose we want to find the intersection between $N$ surfaces where each surface is specified implicitly by a component of vector function $\rates:\Reals^{N+1}\rightarrow\Reals^N$. There are $N$ surfaces, each embedded in $\Reals^{N+1}$. Let's assume that their intersection exists and is not null or degenerate, then their intersection must be a set of one dimensional space curves in an augmented state space $z\in\Reals^{N+1}$ defined by
\begin{align}
    \rates(z) = 0
\end{align}
\begin{figure}[H]
\centering{}
\captionsetup{justification=centering}
\includegraphics[width=5cm]{implicit-surfaces}
\caption{Two implicit surfaces $f_{\theta}(z)=0$ and $g_{\theta}(z)=0$ in $\mathbb{R}^3$ intersecting to form a\\ space curve which is tangent to field $\tangent(z)$ and perpendicular to gradients $\partial_{z}f_{\theta}$ and $\partial_{z}g_{\theta}$}
\label{fig:implicit-surfaces}
\end{figure}

An expression for the field $\tangent(z)$ tangent to the set of curves would allow us to take derivatives and integrals along them. Fortunately the tangent field can be constructed by ensuring it is perpendicular to the gradient $\partial_z$ of each component of $\rates$ as illustrated by an example two component system in Figure \ref{fig:implicit-surfaces}. The tangent field $\tangent(z)$ can be constructed perpendicular to all gradient vectors using the properties of the determinant \cite{Goldman2005CurvatureSurfaces}
\begin{align}
    \tangent(z):=
    \label{eq:tangent-field}
    \left|\begin{matrix}
        \hat{z} \\
        \,\partial_{z}\rates\,
    \end{matrix}\right|
    \qquad\tangent : \Reals^{N+1}\rightarrow\Reals^{N+1}\\
    =\sum_{i=1}^{N+1}\hat{z}_{i}(-1)^{i+1} \left|\frac{\partial \rates}{\partial(z\setminus z_{i}) }\right|
\end{align}
where $\hat{z}$ is a collection of unit basis vectors in the $\Reals^{N+1}$ space and $\partial_{z}\rates$ is an $N\times(N+1)$ rectangular Jacobian matrix of partial derivatives and $z\setminus z_{i}$ denotes the $N$ dimensional vector $z$ with component $z_{i}$ removed. This construction ensures that the dot product of this field any gradient of a component of $\rates$
\begin{align}
    \tangent(z)\cdot\partial_z f_{\theta} =
    \left|\begin{matrix}
        \partial_z f_{\theta} \\
        \,\partial_{z}\rates\,
    \end{matrix}\right|
    \quad =0 \quad\forall f_{\theta}\in \rates
\end{align}
since the determinant of any matrix with two identical rows or columns is zero. Note that the tangent field $\tangent(z)$ is actually defined for all values of $z$ where adjacent field lines trace out other level sets where $\rates(z)\neq0$. Furthermore deformations with respect to $\theta$ are always orthogonal to the tangent
% \todo{numerically true. analytic proof?}
\begin{align}
    \tangent(z)\cdot\frac{d\tangent}{d\theta}=0
\end{align}

\begin{figure}[H]
\centering{}
\captionsetup{justification=centering}
\includegraphics[width=14cm]{determinant-curvature}
\caption{Left/Right : Determinant $\Det(z)$ and tangent field $\tangent(z)$ for the saddle-node/pitchfork models for some set values of $\theta$ revealing that $\Det(z)=0$ defines bifurcations}
\label{fig:determinant-curvature}
\end{figure}

Figure \ref{fig:determinant-curvature} shows how the bifurcation curve defined by $\rates(z)=0$ picks out one of many traces in tangent field $\tangent(z)$ for the saddle and pitchfork. The tangent field $\tangent(z)$ can always be analytically evaluated by taking the determinant in \eqref{eq:tangent-field}. We will proceed with calculations on $\tangent(z)$ in the whole space $z$ and pick out a single trace by solving $\rates(z)=0$ later. For our two models
\begin{align}
    \underset{\mathrm{saddle-node\,\,model}}{
    \tangent(z)=\hat{u}-(\,3\theta_2 u^2+\theta_1\,)\,\hat{p}}
    \qquad\qquad
    \underset{\mathrm{pitchfork\,\,model}}{
    \tangent(z)=u\hat{u}-(\,3\theta_2 u^2+p\,)\,\hat{p}}
    \label{eq:tangent-field-examples}
\end{align}

\subsection{Evaluating Curvature of the Determinant}
Bifurcation points are defined as values of $z$ along the bifurcation curve where eigenvalues $\lambda$ of the Jacobian $\partial_u \rates(z)$ cross either the real or imaginary axis in the complex plane. Restricting our case to $\lambda\in\Reals$ for now, it would be sufficient to look for values of $z$ where one of the eigenvalues $\lambda$ crosses zero. Therefore the determinant of the Jacobian
\begin{align}
    \Det(z) := \left|\,\frac{\partial\rates}{\partial u}\,\right|
\end{align}
crossing zero can be used as a readout of whether a bifurcation has occurred. Figure \ref{fig:determinant-curvature} reveals this is indeed also true for any value $z$ --- not just along $\rates(z)=0$ --- for the saddle-node and pitchfork. The tangent field $\tangent(z)$ only folds when $\Det(z)=0$. Plotting the value of the determinant along $\rates(z)=0$ from Figure \ref{fig:determinant-curvature} would give rise to Figures \ref{fig:saddle-node} and \ref{fig:pitchfork}.

Consequently the curvature of the determinant $\partial_{\tangent}^2\Det$ along the tangent field $\tangent(z)$ in the vicinity of bifurcations tends to increase. Fortunately this curvature can always be obtained in closed analytical form. The second directional derivative with respect to tangent field $\tangent(z)$
\begin{align}
    \partial_{\tangent}^2\Det &:=
    \bigg(
        \partial_z\left(
            \partial_z\Det \cdot \hat{\tangent}
        \right)
    \bigg)\cdot \hat{\tangent}\quad
    \mathrm{where} \quad \hat{\tangent}:=\frac{\tangent}{|\tangent|}\\
    &=
    \hat{\tangent}\cdot\partial_z^2\Det\cdot\hat{\tangent}
    \quad+\quad
    \partial_z\Det\cdot\partial_z\hat{\tangent}\cdot\hat{\tangent}
    % & = \frac{1}{|\tangent|^2}\left(
    %     \tangent\cdot\partial_z^2\Det\cdot\tangent
    %     \quad+\quad
    %     \partial_z\Det\cdot
    %     \left(
    %     \mathbb{I}-\hat{\tangent}\hat{\tangent}^\top
    %     \right)\cdot\partial_z\tangent
    %     \cdot\tangent
    % \right)
\end{align}
The first term in the expression above is the usual result involving the Hessian matrix $\partial_z^2\Det$ when taking second order directional derivatives. The second term involving the gradient $\partial_z\Det$ appears because $\hat{\tangent}(z)$ is a function of $z$ giving rise to a non-zero Jacobian matrix $\partial_z \hat{\tangent}$. Using expressions for tangent fields \eqref{eq:tangent-field-examples} the determinant curvatures for our models become
\begin{align}
    \underset{\mathrm{saddle-node\,\,model}}{
    \partial_{\tangent}^2\Det =
    -\frac{6 \theta_2 \left(\theta _1^2-9 \theta _2^2 u^4+1\right)}
    {\left(\left(\theta _1+3 \theta _2 u^2\right){}^2+1\right){}^2}}
    \qquad
    \underset{\mathrm{pitchfork\,\,model}}{
    \partial_{\tangent}^2\Det =
    -\frac{2 u^2 \left(p+3 \theta _2 u^2\right) \left(9 \theta _2 \left(p+\theta _2 u^2\right)+1\right)}{
    ( \left(p+3 \theta _2 u^2\right)^2+u^2 )^2
    }}
\end{align}

\subsection{Semi-supervised Cost Function}

In order for predicted bifurcations $\predictions(\theta)$ to match targets $\targets$ we need to evaluate some form of error term on the bifurcation curve defined by $\rates(z)=0$. We can see from Figures \ref{fig:saddle-node} and \ref{fig:pitchfork} that predictions can be identified by look for points $p$ along the curve where the determinant $|\frac{\partial\rates}{\partial u}|=0$. We can constrain the cost function to be evaluated in the vicinity of predicted bifurcation points by weighting the likelihood by a function $\omega_\theta(z|\alpha)$ sharply peaked where $|\frac{\partial\rates}{\partial u}|=0$ and going to zero elsewhere. Similarly, we can construct a mixture distribution $\mathcal{N}(p|\targets,\varepsilon)$ peaking at the locations of targets $\targets$ and falling to zero elsewhere. If we choose Gaussian basis functions for both the target mixture and prediction weighting, the likelihood $L_{\theta}(z|\targets)$ takes the form
\begin{align}
    L_{\theta}(z|\targets) := 
    \frac{\omega_\theta(z|\alpha)\,\mathcal{N}(p|\targets,\varepsilon)}{\int_{\rates(z)=0} \omega_\theta(z|\alpha)\,\mathrm{d}z}\label{eq:likelihood}\qquad\qquad\qquad\\
    \mathcal{N}(p|\targets,\varepsilon):=\frac{\sum_{p'\in\targets}\mathbb{e}^{-|p'-p|^2/\varepsilon}}{|\targets|\sqrt{\varepsilon\pi}}
    \quad
    \omega_\theta(z|\alpha):= \mathbb{e}^{-\left|\frac{\partial F_{\theta}}{\partial u}\right|^2/\alpha}
\end{align}
where $\varepsilon$ and $\alpha$ are precision hyper-parameters. Figure \ref{fig:likelihood} shows how the likelihood $L_{\theta}(z|\targets)$ along the curve is indeed maximal when the bifurcations $\predictions(\theta)$ of the saddle-node model match the targets $\targets$.

\begin{figure}[H]
\centering{}
\captionsetup{justification=centering}
\includegraphics[width=7cm]{likelihood-1.pdf}
\includegraphics[width=7cm]{likelihood-2.pdf}
\caption{Likelihood $L_{\theta}(z|\targets)$ shown in dark gold is sharply peaked around the bifurcation points $\predictions(\theta)$ and is maximal when they match the targets (Right). The light gold envelope shows the maxima of the likelihood as the points $\predictions(\theta)$ sweep across different values of $p$}
\label{fig:likelihood}
\end{figure}

>>>>

\input{style}

\title{Inference of Bifurcations\\with Differentiable Continuation}
\author{
        Gregory Szep,\quad Attila Csik\'asz-Nagy\\King's College London\\London, WC2R 2LS\\
        \texttt{gregory.szep@kcl.ac.uk}
    \And
        Neil Dalchau\\Microsoft Research Cambridge\\Cambridge, CB1 2FB\\
        \texttt{ndalchau@microsoft.com}
}

\begin{document}
\input{notation}
\maketitle

\begin{abstract}
    In this work we propose a gradient-based semi-supervised approach for matching target bifurcations with parameterised differential equation models. The cost function contains a supervised term that is minimal when predicted bifurcations match the targets and an unsupervised bias that encourages bifurcations by maximising zero crossings in the determinant of the Jacobian. The calculation of gradients with respect to parameters shares the same computational complexity as deflated pseudo-arclength continuation used to calculate the bifurcation diagram. We demonstrate model synthesis with minimal models which explore the space of saddle-node and pitchfork diagrams, a genetic toggle switch from synthetic biology and the FitzHugh-Nagumo model. Furthermore, the cost landscape allows us to  organise models in terms of topological and geometric equivalence.
\end{abstract}

\section{Introduction}

Backpropagation through differential equation solves has been a breakthrough over the past couple of years \cite{Chen2018NeuralEquations,Rackauckas2019DiffEqFlux.jl-AEquations} that enabled scalable parameter inference for differential equations. Determining what model parameters should be for a given set of observations, in the setting of biology and engineering, are known as inverse problems \cite{Abdulla2009InverseBiology}.

Optimisation targets, however, have mostly been expressed in the spatio-temporal domain. Acquisition of such data can be costly and can often contain over-constraining information generated by processing steps or the measurement device rather than the state variables that underpin the observed mechanism. In microscopy, for example, data is often reported in arbitrary fluoresce units allowing the observer to shift and scale data arbitrarily. Furthermore, such data may also not contain sufficient information about dynamical transients in order to identify kinetic parameters. The emerging picture suggests that identification of the qualitative behaviour -- the bifurcation diagram -- should precede any attempt at inferring kinetic parameters \cite{Stumpf2019ParameterBifurcations}. Techniques for back-propagating through implicit equation solvers have also been developed \cite{Look2020DifferentiableLayers,Bai2019DeepModels} although to the best of the authors' knowledge have not been applied to bifurcation diagrams at the time of writing this paper.

In this work we propose a gradient-based semi-supervised approach that focuses on fitting high-level qualitative constraints, defined by state space structures, rather than kinetics. Drawing inspiration from implicit solvers \cite{Look2020DifferentiableLayers,Bai2019DeepModels} to calculate gradients we find that their computation shares the same  complexity as the algorithm used to calculate the bifurcation diagram. We use a predictor-corrector method called deflated pseudo-arclength continuation \cite{Farrell2016TheDiagrams,Veltz2019PseudoArcLengthContinuation.jl}, originally developed for partial differential equations. In the case of partial differential equations the computational complexity of calculating a single bifurcation diagram is not bounded since superpositions of localised solutions may give rise to uncountably many branches \cite{Avitabile2010ToEquation}. The complexity of computing a single branch, however, is bounded by the complexities of the chosen eigenvalue solver and corrector, and further decreased by adaptive stepping procedures \cite{Aruliah2016AlgorithmContinuation}.

We find that the cost function landscape contains basins that not only allow us to synthesise models with a desired bifurcation structure but also allow us to organise models in terms of topological and geometric equivalence. We discuss the relevance of this in model selection. In summary, our paper has the following main contributions:

\begin{itemize}
    \item Defined a differentiable semi-supervised cost function for encouraging and placing co-dimension one bifurcations to user-specified target locations
    \item Implementation of method in Julia package \texttt{FluxContinuation.jl} leveraging automatic differentiation tooling in \texttt{Flux.jl}
    \item Leveraging the cost landscape for a novel way of organising differential equation models in terms of geometric and topological equivalence
\end{itemize}

\subsection{Related Works | \textit{in progress}}

\paragraph{Bifurcation control} and \textit{inverse} bifurcation analysis literature applies control theory to enable design and identification of dynamical systems with qualitative constraints. The first works involved the calculation of the minimal distances to bifurcation manifolds \cite{Iwasaki1997AnType,Lu2006InverseSystems,Dobson2004DistanceBifurcations}. Those optimisations did not involve closed form expressions and the challenge of finding the manifolds to being with was left to random sampling.

\paragraph{Matching vector fields} From a dynamical systems perspective a given bifurcation curve constrains a state-vector field geometry. A set of target bifurcations may also be represented as a vector field flow and thus inferring bifurcations becomes a problem in matching flows in vector sub-spaces. Note that this would involve matching direction but not the magnitude of the flow.

\begin{itemize}
    \item smooth and match estimator methods \cite{Ranciati2017BayesianParameters}
    \item robots learning limit cycles \cite{Khadivar2021LearningBifurcations}
\end{itemize}

\paragraph{Non-differentiable methods} 
A large body of work is dedicated to finding conditions for fold-bifurcations, multistability and other

\begin{itemize}
    \item bifurcation inference using Mixed Integer Nonlinear Programming \cite{Otero-Muras2018Optimization-basedModels}
    \item Random sampling and genetic algorithms \cite{Chickarmane2005BifurcationTool,Conrad2006BifurcationClock}
\end{itemize}

\subsection{Preliminaries}
Suppose we parameterise a set of differential equations for states $u\in\Reals^N$ with a function $\rates$ in an unknown parameter space $\theta\in\Reals^M$. We would like these differential equations to obey a set of target bifurcations $\targets:=\{p_1\dots p_K\}$ along a known bifurcation parameter $p\in\Reals$. Let the differential equations be defined as
\begin{align}
	\partial_t u=\rates(z)
	\qquad\mathrm{where}\quad z:=(u,p)\quad
	\rates : \Reals^{N+1}\rightarrow\Reals^N
	\label{eq:model}
\end{align}
For a given set of parameters $\theta$ one could compute the set of predicted bifurcations $\predictions(\theta)$ using parameter continuation methods \cite{Veltz2019PseudoArcLengthContinuation.jl,Farrell2016TheDiagrams}. Our goal is to find optimal parameters $\theta^*$ that match predictions $\predictions(\theta^*)$ to specified targets $\targets$. We must design a suitable cost function $\loss$ so that
\begin{align}
    \theta^* := \mathrm{argmin}_{\theta} \loss(\theta|\targets)
    \label{eq:optimal-theta}
\end{align}
The optimal $\theta^*$ is not expected to always be unique, but is in general a manifold representing the space of qualitatively equivalent models. The cost function $\loss$ must have some distance measure between the two sets $\predictions(\theta)$ and $\targets$ with a bias that encourages equivalent cardinality $|\predictions(\theta)|=|\targets|$. This is especially important in the case where there are no predictions $|\predictions(\theta)|=0$.

For clarity, we guide the reader through the methods with the following minimal models that explore the space of saddle-nodes $\rates(z) = p + \theta_{1}u+\theta_{2}u^3$ and pitchforks $\rates(z) = \theta_{1} + p u+\theta_{2}u^3$. These minimal models are shown with targets $\targets$ in Figure \ref{fig:saddle-node} and Figure \ref{fig:pitchfork} respectively. These figures show that the determinant $\Det=0$ at the bifurcations and that its directional derivative along the bifurcation curve $\frac{d}{ds}\Det$ is finite and non-zero, at least in the non-degenerate case. This gives us our first hint of what should be optimised to increase the number of predictions $|\predictions(\theta)|$.

\begin{figure}
\centering
\includegraphics[width=11cm]{saddle-node}
\caption{Saddle-node model $\rates(z) = p + \theta_{1}u+\theta_{2}u^3$ with $\theta=(5/2,-1)$ and targets $\targets=\{-1,1\}$. Determinant in red. Lighter shades indicate the determinant crossing zero for unstable solutions}
\label{fig:saddle-node}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=11cm]{pitchfork}
\caption{Pitchfork model $\rates(z) = \theta_{1} + p u+\theta_{2}u^3$ with $\theta=(1/2,-1)$ and target $\targets=\{0\}$. Determinant in red. Lighter shades indicate the determinant crossing zero for unstable solutions}
\label{fig:pitchfork}
\end{figure}

\section{Proposed Method}
\subsection{Semi-supervised Cost Function}

\begin{figure}
    \centering
    \includegraphics[width=10cm]{bifurcation-measure}
    \caption{Bifurcation measure $\measure(z)$ shown as a gold thickness envelope. The thicker the envelope the larger the value of the measure at that location in the curve. Three curves are shown for three different values $\theta$, demonstrating how maximising $\measure(z)$ along the curve encourages new bifurcations}
    \label{fig:bifurcation-measure}
\end{figure}

In order for predicted bifurcations $p\in\predictions(\theta)$ to match targets $p'\in\targets$ we need to evaluate some form of error term $|p-p'|$ on the bifurcation curve defined by $\rates(z)=0$. We expect the supervised term to be some mean over the targets $\targets$ and predictions $\predictions(\theta)$. We choose a geometric mean over the predictions and an arithmetic mean over targets to account for cases where $|\predictions|\neq|\targets|$ and discourage multiple predictions matching the same target in cases where $|\predictions|\geq|\targets|$. This term is equal to zero when all targets are matched by at least one prediction
\begin{align}
    \frac{1}{|\targets|}\sum_{p'\in\targets}
    \left(\prod_{p\in\predictions(\theta)}|p-p'|\right)
    ^{\frac{1}{|\predictions|}}
    \label{eq:supervised-term}
\end{align}
We can see from Figures \ref{fig:saddle-node} and \ref{fig:pitchfork} that predictions $\predictions(\theta)$ can be identified by looking for points $p$ along the curve where the determinant $\Det=0$. Meanwhile the directional derivative along the bifurcation curve is finite $\frac{d}{ds}\Det\neq 0$ in the non-degenerate case. Using these quantities we can define a positive semi-definite measure $\measure(z)$ of zero crossings in the determinant along the bifurcation curve; as shown in Figure \ref{fig:bifurcation-measure} this can be used as a proxy for the probability of a bifurcation at location $p$ for a given set of parameters $\theta$. The total bifurcation measure $\Psi(\theta)$ is written
\begin{align}
    \Psi(\theta):=\frac{
        \int_{\rates(z)=0}\!
        \measure(z)\,\mathrm{d}z
    }{
        \int_{\rates(z)=0}\!\
        \mathrm{d}z
    }
    \quad\mathrm{where}\quad
    \measure(z):=
    \left(1+\left|\frac{\Det}{\frac{d}{ds}\Det}\right|\right)^{-1}
    \label{eq:measure}
\end{align}
Here we denote $\int_{\rates(z)=0}\mathrm{d}z$ as the sum of the line integrals in $z\in\Reals^{N+1}$ defined by the steady states $\rates(z)=0$. The measure $\measure(z)=1$ at bifurcation points and goes to zero an odd number of times between bifurcations. This is because $\Det$ must eventually turn around in order to return back to zero, resulting in $\frac{d}{ds}\Det\rightarrow0$ and hence $\measure(z)\rightarrow0$ for each turning point. As $\Det$ diverges we approach regimes far away from any closest bifurcation and hence $\measure(z)\rightarrow0$. We would still like to have non-zero gradients with respect to $\theta$ in this regime, and therefore $\measure(z)$ was designed to go to zero slowly. The total measure $\Psi(\theta)$ is normalised such that $\Psi(\theta)\rightarrow1$ in the regimes where the observed parameter region $p$ is densely packed with bifurcations. In Appendix \ref{appendix:measure} we see that $\measure(z)$ is also invariant under certain classes of transformations on $\rates(z)$ which do not change bifurcation locations. The measure is added to the supervised term as if it were a likelihood; this defines the semi-supervised cost function as
\begin{align}
    \loss(\theta|\targets):=\frac{1}{|\targets|}\sum_{p'\in\targets}
    \left(\prod_{p\in\predictions(\theta)}|p-p'|\right)
    ^{\frac{1}{|\predictions|}}\!\!\!
    -\big(|\targets|-|\predictions|\big)\,\,\lambda \log\Psi(\theta)
    \label{eq:cost}
\end{align}
The pre-factor $|\targets|-|\predictions|$ in the unsupervised term ensures that the gradients are always pushing optimisers towards a state where $|\targets|=|\predictions|$. This can be seen as a step-wise annealing of the unsupervised term until the desired state is reached. Note that in principle while $\predictions(\theta)$ depends smoothly on $\theta$ the number of predictions $|\predictions|$ does not. The gradient $\partial_{\theta}|\predictions|=0$ everywhere and is an infinite impulse at locations where the number of predictions changes. In practice optimisers are unlikely to reach those locations and we can safely drop the dependency on $\theta$. The supervised term must be re-written in such a way so that we may explicitly differentiate the cost with respect to $\theta$. This can be done in the following way
\begin{align}
    \frac{1}{|\targets|}\sum_{p'\in\targets}
    \left(\prod_{p\in\predictions(\theta)}|p-p'|\right)
    ^{\frac{1}{|\predictions|}}\!\!=\,\,
    \frac{1}{|\targets|}\sum_{p'\in\targets}
    \exp\left[
    \frac{1}{|\predictions|}\int_{\rates(z)=0\,\Det=0}
    \!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!
    \mathrm{d}z\quad
    \log|p-p'|
    \right]
    \label{eq:supervised-rewrite}
\end{align}
In effect we have replaced the set representation $\predictions(\theta)$ with a line integral over the bifurcation curve $\rates(z)=0$ subject to constraint $\Det=0$. In Figure \ref{fig:determinant-field} we can indeed see that bifurcations $\predictions(\theta)$ are picked out by intersections between the two level sets. Now that the cost function is written in terms of line integrals $\int_{\rates(z)=0}\mathrm{d}z$ we need to know how to differentiate them with respect to $\theta$, differentiate along them as in equation \eqref{eq:measure} and know how to deal with constraint $\Det=0$. The following section shows how this can be done with a vector field representation of the bifurcation curve.

\subsection{Bifurcation Curves as Tangent Fields}
\label{section:tangent-fields}

Let each component of the vector function $\rates$ in the model \eqref{eq:model} implicitly define a surface embedded in $\Reals^{N+1}$. Let's assume that the intersection of these $N$ surfaces exists and is not null or degenerate, then the steady states of \eqref{eq:model} must be a set of one dimensional space curves in $z\in\Reals^{N+1}$ defined by
\begin{align}
    \rates(z) = 0
\end{align}
An expression for the field $\tangent(z)$ tangent to the set of curves would allow us to take derivatives and integrals along the bifurcation curve. This is exactly what we need to do to evaluate our cost function \ref{eq:cost}. Fortunately the tangent field can be constructed by ensuring it is perpendicular to the gradient $\partial_z$ of each component of $\rates$ as illustrated by an example two component system in Figure \ref{fig:implicit-surfaces}. The tangent field $\tangent(z)$ can be constructed perpendicular to all gradient vectors using the properties of the determinant \cite{Goldman2005CurvatureSurfaces}
\begin{align}
    \tangent(z):=
    \label{eq:tangent-field}
    \left|\begin{matrix}
        \hat{z} \\
        \,\partial_{z}\rates\,
    \end{matrix}\right|
    \qquad\tangent : \Reals^{N+1}\rightarrow\Reals^{N+1}\\
    =\sum_{i=1}^{N+1}\hat{z}_{i}(-1)^{i+1} \left|\frac{\partial \rates}{\partial(z\setminus z_{i}) }\right|
\end{align}
where $\hat{z}$ is a collection of unit basis vectors in the $\Reals^{N+1}$ space and $\partial_{z}\rates$ is an $N\times(N+1)$ rectangular Jacobian matrix of partial derivatives and $z\setminus z_{i}$ denotes the $N$ dimensional vector $z$ with component $z_{i}$ removed. This construction ensures perpendicularity to any gradients of $\rates$
\begin{align}
    \tangent(z)\cdot\partial_z f_{\theta} =
    \left|\begin{matrix}
        \partial_z f_{\theta} \\
        \,\partial_{z}\rates\,
    \end{matrix}\right|
    \quad =0 \quad\forall f_{\theta}\in \rates
\end{align}
since the determinant of any matrix with two identical rows or columns is zero. Note that the tangent field $\tangent(z)$ is actually defined for all values of $z$ where adjacent field lines trace out other level sets where $\rates(z)\neq0$. Furthermore deformations with respect to $\theta$ are always orthogonal to the tangent
\begin{align} % \todo{numerically true. analytic proof?}
    \tangent(z)\cdot\frac{d\tangent}{d\theta}=0
\end{align}
\begin{figure}
\centering
\includegraphics[width=13cm]{determinant-field}
\caption{Left/Right : Determinant $\Det$ and tangent field $\tangent(z)$ for the saddle-node/pitchfork models for some set values of $\theta$ revealing that $\Det=0$ defines bifurcations}
\label{fig:determinant-field}
\end{figure}
Figure \ref{fig:determinant-field} shows how the bifurcation curve defined by $\rates(z)=0$ picks out one of many level sets or traces in tangent field $\tangent(z)$ for the saddle and pitchfork. The tangent field $\tangent(z)$ can always be analytically evaluated by taking the determinant in \eqref{eq:tangent-field}. We will proceed with calculations on $\tangent(z)$ in the whole space $z$ and pick out a single trace by solving $\rates(z)=0$ later. For our two models
\begin{align}
    \underset{\mathrm{saddle-node\,\,model}}{
    \tangent(z)=\hat{u}-(\,3\theta_2 u^2+\theta_1\,)\,\hat{p}}
    \qquad\qquad
    \underset{\mathrm{pitchfork\,\,model}}{
    \tangent(z)=u\hat{u}-(\,3\theta_2 u^2+p\,)\,\hat{p}}
    \label{eq:tangent-field-examples}
\end{align}
Figure \ref{fig:determinant-field} reveals that $\Det=0$ is also a level set and that the intersection with level set $\rates(z)=0$ defines the bifurcations at specific parameter $\theta$. In this particular setting we can see that the tangent field $\tangent(z)$ only folds when $\Det=0$. Plotting the value of the determinant along $\rates(z)=0$ from Figure \ref{fig:determinant-field} would give rise to Figures \ref{fig:saddle-node} and \ref{fig:pitchfork}. The directional derivative of the determinant $\Det$ along the tangent field $\tangent(z)$ is defined as
\begin{align}
    \frac{d}{ds}\Det := \hat{\tangent}(z) \cdot \frac{\partial}{\partial z}\Det 
\end{align}
where $\hat{\tangent}(z)$ is the unit tangent field. Finally, by applying the generalised Leibniz integral rule, it is also possible to use $\tangent(z)$ to evaluate gradients of line integrals  $\int_{\rates(z)=0}\mathrm{d}z$. In a similar vein to back-propagation through neural differential equations \cite{Chen2018NeuralEquations} we would like to be able to differentiate the cost $\loss(\theta|\targets)$ without having to differentiate through the operations of the solver that finds $\rates(z)=0$. Fortunately this is all possible if we apply a similar strategy to that used by implicit layers \cite{Look2020DifferentiableLayers}; see Appendix \ref{appendix:space-curve} and \ref{appendix:deformation} for derivations.
